# (APPENDIX) Appendix {-}

```{r warning = FALSE, message = FALSE, echo = FALSE}
library(tidyverse)
library(ggpubr)
library(here)
library(rstatix)
library(effectsize)
library(car)
```

# Appendix A: Technical details of `anova()` 

```{r echo = FALSE}
w10_flow <- read_csv(here("data", "week_10", "w10_flow.csv"))
flow_block1 <- lm(DFS_Total ~ GoldMSI, data = w10_flow)
flow_block2 <- lm(DFS_Total ~ GoldMSI + openness, data = w10_flow)
```

Technical note: this test works not too unlike a regular ANOVA, except the F-test is being conducted on the residual sums of scores. From a mathematical point of view, we are essentially conducting an F-test on the change in the residual sum of squares with the following formula:

$$
F(df_{df_b - df_a}, df_a) = \frac{MS_{comp}}{MS_{a}} = \frac{(SS_b - SS_a)/(p_a - p_b)}{SS_a/df_a}
$$

Consider two models, Model A and Model B. Imagine Model B is a nested version of Model A - i.e. it it the same model as Model A but with less predictors. In our case, imagine Model B is `flow_block1` (which only had one predictor) and Model A is `flow_block2` (which had two). $p_a$ is the number of coefficients in Model A *including* the intercept, and same with $p_b$. 

The exact process is:

1.    Calculate the difference between residual SS in the two models - this is the $(SS_b - SS_a)$ part of the formula above. This is just the difference in RSS between model 1 (model B) and 2 (model A), i.e. 10124.2 - 9982.9 = 141.3.

2.    Calculate the difference in df $(p_a - p_b)$. In `flow_block1` we have one predictor and one intercept, so we have 2 terms - this is $p_b$. In `flow_block2` we have two predictors and one intercept, which makes $p_a = 3$. Therefore, $(p_a - p_b) = 3 - 2 = 1$.

3.    Calculate a mean square ratio for the comparison, which is $MS_{comp}$. Essentially, we divide the result in step 1 (143.1) by the result in step 2 (1). This follows the same formula for mean squarews as we have seen before: $MS = \frac{SS_{comp}}{df_{comp}}$, so 
$MS_{comp} = \frac{141.3}{1} = 141.3.$ While this is identical to the sum of squares value in the table above, note that this is *not* the same value.

4.    Calculate a mean square ratio between RSS and df for the new model. This is the $SS_a/df_a$ part of the equation. $df_a$ is calculated as $n - p_a$, where n is the original sample size. So $df_a = 811 - 3 = 808$. Note that the value for row 2 (which corresponds to Model A/`flow_block2`) under `Res.df` is 808. 

Note that $df_b$ is the same; $n - p_b = 811 - 2 = 809$.

Same deal as above after that, except this time we use the values from the new model only, i.e. residual SS for model A (`flow_block2`) and the residual df. 

$MS_a = \frac{SS_{a}}{df_{a}}$
$MS_a = \frac{9982.9}{808} = 12.33507$

5.    Calculate an F ratio between the MS of the comparison and the MS of the new model to calculate a value for F.

This is exactly the same formula as it would be for a regular ANOVA, just that now we are doing:

$$
F = \frac{MS_{comp}}{MS_{a}}
$$

$F = \frac{141.3}{12.33507} = 11.45514$ 

6. Calculate a p-value for this F-statistic by comparing the p against an F distribution. The two dfs in the original formula are a) $df_b - df_a$ and b) $df_a$. Which means:

-   $df_b$ is 809 - $df_a$ is 808 = 1
-   $df_a$ is 808

So we end up with a test statistic of $F(1, 808) = 11.455$. We can use this to calculate a p-value by calculating the probability of getting a value of at least 11.455, on an F distribution with degrees of freedom parameters described above. We can visualise this below. Note that because the values for F are so infinitesimally small with these parameters and at this F-value, I've zoomed in the plot to visualise the highlighted area:

```{r}
tibble(
  x = seq(0, 15, by = .5),
  y = df(x, df1 = 1, df2 = 809)
) %>%
  ggplot(
    aes(x = x, y = y)
  ) +
  geom_line(linewidth = 1) +
  theme_pubr() +
  geom_vline(xintercept = 11.455, linewidth = 1, colour = "royalblue") +
  annotate("text", x = 12, y = 0.0024, label = "F = 11.455") +
  stat_function(fun = df, args = list(df1 = 1, df2 = 809), 
                geom = "area", xlim = c(11.455, 15),
                fill = "royalblue", alpha = 0.5) + 
  scale_x_continuous(expand = c(0, 0), limits = c(10, NA)) + 
  scale_y_continuous(expand = c(0, 0), limits = c(0, 0.003)) +
  labs(x = "F-value", y = "Density")

```

R can manually calculate a p-value with the `pf()` function. `pf()` will calculate the probability of a value on the F distribution, given the two degrees of freedom parameters to characterise the distribution. `lower.tail = FALSE` is used to indicate that we want to calculate the probability of getting something *above* our critical F-value; `lower.tail = TRUE` would calculuate the probability *below* it.

```{r}
pf(11.45514, df1 = 1, df2 = 808, lower.tail = FALSE)
```


Note that our p-value isn't exactly the same as the value in the table - this is because we've used rounded values. The code below extracts the unrounded values and uses them in the calculations. As you can see we get the exact p-value in the table.

```{r}
x <- anova(flow_block1, flow_block2)
x

# Using the output from anova() to manually calculate p-value
SSb <- x$RSS[1]
SSa <- x$RSS[2]
pa <- length(coef(flow_block2))
pb <- length(coef(flow_block1))
dfa <- nrow(w10_flow) - pa
dfb <- nrow(w10_flow) - pb

f_val <- ((SSb-SSa)/(pa-pb))/(SSa/dfa)

pf(f_val, df1 = dfb-dfa, df2 = dfa, lower.tail = FALSE)

```

