# Preface {.unnumbered}

This is an R adaptation of the quantitative research/statistics content from a subject called Research Process for Music Psychologists (MUSI90252). The subject is an introduction to research methods for new Masters and PhD students in music psychology and music science at the University of Melbourne.

**Some context, and why this book exists**

The subject's statistics material was originally written for [Jamovi](https://www.jamovi.org/), given its ease of use and impressive functionality. The reason why we (the original subject coordinator and I) preferenced Jamovi in the first instance is because many of our students do not come from psychological science, statistics or research backgrounds, so statistical analysis is often an entirely new frontier for them. Jamovi provides a *free* point-and-click interface that will be familiar to users of SPSS and other platforms who may be used to a point-and-click approach. I actually think this is a great thing, and I believe one of Jamovi's greatest strengths is how easy it is to use. 

However, I have also strongly encouraged any music psychology students doing quantitatively-oriented research to consider learning R in the long term. R is therefore also offered as an option for this subject, and this book is meant to serve as the R-specific version of the subject material.

By and large the content is a very faithful reproduction of the subject's original Jamovi-oriented content, with the following exceptions:

- Some additional commentary has been added for R-specific material, such as information on certain functions.
- Some embedded content in the Canvas version is not available here (mainly Jamovi-specific content, and Readings Online content).
- Some content has been re-organised because of how R outputs things compared to Jamovi. 
- The first chapter is a very (*very*) brief overview of how to use core R and `tidyverse` functions.

Also like the Jamovi version, **Parts I and II** of this book form the relevant assessable content (Weeks 5-10), while Part III is an extension of the material beyond the core subject. It is also provided here though for interested readers.

Where possible, the subject content has been written to achieve **parity with Jamovi** first and foremost, to reduce discrepancies and friction between the two versions of the subject material. The material in this book has been designed to align with what Jamovi provides as closely as possible, in terms of both functionality and output. As an avid R user myself though, I am aware that this means that there will be things that R can do that we do not closely discuss, or there may be things that Jamovi does that are somewhat different to what other programs (or R packages) may do or recommend. R can and will do so much more than what is presented in this book, and I strongly encourage interested students to seek out this content separately.

**Information boxes**

Throughout the book you will see a number of **boxes**, highlighted with different colours and labelled with different icons and labels (one of the major upgrades to this book as a result of moving to Quarto). Some of them will be hidden by default, and can be expanded by clicking on them. Here is what the various boxes mean:

::: {.callout-note title="Information boxes"}
These blue info boxes will contain additional supplementary information about some of the concepts discussed in the subject/book. 
:::

::: {.callout-tip title="R Notes"}
Throughout the book there are various *R Notes*, which provide additional tips, clarifications or information about functions or procedures specific to R. You will find them in these cute green boxes.
:::

::: {.callout-warning title="Warning boxes"}
These warning boxes indicate topics, concepts or procedures that are important to know/remember, or highlight things to avoid.
:::

**About this book**

Up until 22nd May 2026, this book was written using [bookdown](https://bookdown.org/). I changed the book to [Quarto](https://quarto.org/), partly because of the additional functinality built in to Quarto and partly because I really like how Quarto books look. It has come at the small cost of one or two features - such as the ability to easily split chapters into multiple pages - but overall makes for (imo) a much nicer reading experience.

The bookdown version is available on a separate branch on my GitHub.

Several elements in this new version of the book will be updated as time goes on. This first render in Quarto may be a bit rough around the edges while I work through these.


```{r, eval=FALSE, echo = FALSE}
# bookdown::render_book()
```


```{r eval = FALSE, echo = FALSE}
# install.packages(c(
#   "tidyverse", "here", "psych", "rstatix", "ggpubr", "DescTools", "ggcorrplot", "afex", "emmeans", "effectsize", "lm.beta", "datarium"
))
```


```{r eval=FALSE, echo = FALSE}
# bookdown::serve_book()
```


```{r include=FALSE}
# automatically create a bib database for R packages
# knitr::write_bib(c(
#   .packages(), 'bookdown', 'knitr', 'rmarkdown'
# ), 'packages.bib')
```
